---
title: "Triggerfish_disparity"
date: "Oct 4, 2021"
output: 
  html_document:
      toc: TRUE
      theme:
        bootswatch: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part I: Data and library load  
  
Reading in files and estimating disparity of single value traits using method of Cooney et al. 2017  
  
To begin we will read in data and then format for downstream functions

```{r}
library(phytools)
library(geiger)
library(msir)
traits<-read.csv("~/Documents/Triggerfish_2021/Archiving/All_final_data.csv")
tree<-read.tree("~/Documents/Triggerfish_2021/Archiving/Final_Inform_partitioned.phy")

######################################
##change input $trait per trait Adult
######################################
ADar<-traits$Adult_dorsal_incidence_angle

######################################
##change input $trait per trait juvenile
######################################
JDar<-traits$Juvi_dorsal_incidence_angle

######################################
#Do NOT change below this line if you are not changing the object names
######################################
## Adult trait 1 data prep
ADar<-data.frame(ADar)
row.names(ADar)<-traits$species #make species rownames
#check
#row.names(ADar)
traitoverlap<-name.check(tree, ADar)
#check
#traitoverlap$tree_not_data
triggerTree<-drop.tip(tree, traitoverlap$tree_not_data)
#plot(triggerTree)
## Juvenile trait 1 data prep
######################################
JDar<-data.frame(JDar)
row.names(JDar)<-traits[,1] #make species rownames
#check
#row.names(JDar)
JDar<-na.omit(JDar)
traitoverlap<-name.check(tree, JDar)
#check
#traitoverlap$tree_not_data
#drop
triggerTreeJ<-drop.tip(tree, traitoverlap$tree_not_data)
#plot(triggerTreeJ)
######################################
## get rid of missing values 
######################################
trait1<-na.omit(ADar)
trait1.j<-na.omit(JDar)
newtrait<-trait1[,1]
newtraitj<-trait1.j[,1]
names(newtrait)<-rownames(trait1)
names(newtraitj)<-rownames(trait1.j)

```

## Part II: Core functions  
  
Here are the core functions used downstream, these just need to be read in. In the interest of keeping the markdown less busy, they are not displayed in the html doc. To toggle them to be displayed, change echo=FALSE, to echo=TRUE.  
  
```{r, echo=FALSE}
######################################
##core function
######################################

disparity<-function(tree,trait, startv=0, stopv=14, interval=1)
{
  bt<-branching.times(tree)
  nullfit<-fastAnc(tree, trait,vars=TRUE,CI=TRUE)
  traitvariance<-NULL
  vals<-seq(from=startv, to=stopv, by=interval)
  for (i in 1:length(vals)){
    temp<-names(bt[bt>vals[i]])
    traitvariance[i]<-var(nullfit$ace[temp])
  }
  traitvariance[is.na(traitvariance)] <- 0
  scaleddis<-traitvariance/max(traitvariance)
  return(scaleddis)
}

######################################
##Helper functions to parse null distribution
######################################

get.median<-function(null.dis){
  meds<-matrix(ncol=length(null.dis[1,]))
  for (i in 1:length(null.dis[1,])){
    median(null.dis[,i])->meds[i]
  }
  return(meds)
}
get.low<-function(null.dis){
  lowmeds<-matrix(ncol=length(null.dis[1,]))
  for (i in 1:length(null.dis[1,])){
    min(null.dis[,i])->lowmeds[i]
  }
  return(lowmeds)
}
get.high<-function(null.dis){
  highmeds<-matrix(ncol=length(null.dis[1,]))
  for (i in 1:length(null.dis[1,])){
    max(null.dis[,i])->highmeds[i]
  }
  return(highmeds)
}

######################################
##Plotting functions 
######################################

#please note the defaults are specific to this markdown

twoStageDisparity<-function(tree1=triggerTree, tree2=triggerTreeJ,trait1=trait1,trait2=trait1.j, null.dis,stopv=14, type="accumulation"){
  yy4<-disparity(tree1,trait1[,1])
  yy5<-disparity(tree2,trait2[,1])
  y<-get.median(null.dis)
  y1<-get.low(null.dis)
  y2<-get.high(null.dis)
  ##null
  nulldistro<-loess.sd(seq(0:stopv), y )
  ##adult
  adultdistro<-loess.sd(seq(0:stopv), yy4 )
  ##Juvenile
  juvidistro<-loess.sd(seq(0:stopv), yy5)
  if(type=="accumulation"){
    ######################################
    ##  plot the smoothed lines
    ######################################
   
    #plotting adults null and juveniles
    #null
    plot(nulldistro$x,nulldistro$y,xlim=rev(c(3,15)), type="l")
    lines(rev(nulldistro$x), rev(nulldistro$upper), lty=2, col="gray")
    lines(rev(nulldistro$x), rev(nulldistro$lower), lty=2, col="gray")
    #adult
    lines(rev(adultdistro$x), rev(adultdistro$y), col="blue")
    lines(rev(adultdistro$x), rev(adultdistro$upper), lty=2, col="blue")
    lines(rev(adultdistro$x), rev(adultdistro$lower), lty=2,col="blue")
    #juvenile
    lines(rev(juvidistro$x), rev(juvidistro$y), col="green")
    lines(rev(juvidistro$x), rev(juvidistro$upper), lty=2, col="green")
    lines(rev(juvidistro$x), rev(juvidistro$lower), lty=2,col="green")
  }else {
    ###
    ######################################
    ## niche packing or filling plots
    ######################################
    ###
    adulty<-(adultdistro$y/adultdistro$x)-(nulldistro$y/nulldistro$x)
    adult_upper<-(adultdistro$upper/adultdistro$x)-(nulldistro$upper/nulldistro$x)
    adult_lower<-(adultdistro$lower/adultdistro$x)-(nulldistro$lower/nulldistro$x)
    juviy<-(juvidistro$y/juvidistro$x)-(nulldistro$y/nulldistro$x)
    juviy_upper<-(juvidistro$upper/juvidistro$x)-(nulldistro$upper/nulldistro$x)
    juviy_lower<-(juvidistro$lower/juvidistro$x)-(nulldistro$lower/nulldistro$x)
    #plot
    plot(seq(from=0, to=stopv, by=1),adulty,ylim=c(min(c(adult_lower,juviy_lower)),max(c(adult_upper,juviy_upper))),xlim=rev(c(3,stopv)),type="l",col="blue")
    lines(seq(from=0, to=stopv, by=1),adult_upper, type="l",lty=2,col="blue")
    lines(seq(from=0, to=stopv, by=1),adult_lower, type="l",lty=2,col="blue")

    lines(seq(0:stopv),juviy_upper, type="l",lty=2,col="green")
    lines(seq(0:stopv),juviy, type="l",col="green")
    lines(seq(0:stopv),juviy_lower, type="l",lty=2,col="green")
    abline(h=0)
    
  }  
}

```

## Part III: Null distribution  

This is a quick loop to generate a null distribution using the above functions  

```{r}

######################################
##get the null distribution
######################################

##Change '1000' in two places below to the number of sims you want

null.dis<-matrix(nrow=100,ncol=15)
for (i in 1:100){
  x <- as.data.frame(fastBM(triggerTree))
  #row.names(ADar)
  disparity(triggerTree,x[,1])->null.dis[i,]
}

```

## Part IV: empirical disparity

```{r}
######################################
## get empirical disparity 
######################################

#this is part of a function above but placing it here for verification
#yy4<-disparity(triggerTree,trait1[,1])
#yy5<-disparity(triggerTreeJ,trait1.j[,1])

######################################
## Plotting
######################################

#First plot the accumulation of total disparity over time
twoStageDisparity(tree1=triggerTree, tree2=triggerTreeJ,trait1=trait1,trait2=trait1.j,stopv=14, null.dis, type="accumulation")

#then plot how morphospaces expand or fill
twoStageDisparity(tree1=triggerTree, tree2=triggerTreeJ,trait1=trait1,trait2=trait1.j,stopv=14, null.dis, type="other")

#to get dtt plot 
dttOut<-dtt(triggerTreeJ, JDar, nsim=1000, CI=0.90)
dttOut<-dtt(triggerTree, ADar, nsim=1000, CI=0.90)

```

Repeat the above logic for any traits or loop through the trait names to generate a series of plots, etc. 